---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
<br>
<br>
Below is the code for reading the input data and aggregating by daily activity:
<br>

```{r}
raw_data <- read.csv("../../peer1/activity.csv", header = TRUE, sep = ",")

# Aggregate the data by daily activity
aggregate_data <- aggregate(steps ~ date, raw_data, as.vector, na.action = na.pass)
aggregate_data[, "total_daily"] <- 0
for (i in 1:nrow(aggregate_data)) {
  aggregate_data[i,]$total_daily <- sum(aggregate_data[i,]$steps)
}
```
## What is mean total number of steps taken per day?
<br>
<br>
Histogram of the total number of steps taken each day:

```{r}
hist(aggregate_data$total_daily, xlab = "Total Daily Steps", main = "Histogram of total daily steps")

```

Mean daily steps: `r format(mean(aggregate_data$total_daily, na.rm = TRUE), digits = 2)`
<br>
Median daily steps: `r format(median(aggregate_data$total_daily, na.rm = TRUE), digits = 2)`

## What is the average daily activity pattern?

Below is the code for aggregating the data by interval and plotting the average number of steps for an interval:

```{r}

#Aggregate by interval
aggregate_data2 <- aggregate(steps ~ interval, raw_data, as.vector, na.action = na.pass)

aggregate_data2[, "average_by_interval"] <- 0
for (i in 1:nrow(aggregate_data2)) {
  aggregate_data2[i,]$average_by_interval <- mean(aggregate_data2[i,]$steps, na.rm = TRUE)
}

plot(aggregate_data2$interval, aggregate_data2$average_by_interval, ylim = c(0, 250), type = "l", main = "Average steps by interval", xlab = "Interval", ylab = "Average steps")


```
<br>
On average, maximum number of steps are taken in interval: `r aggregate_data2[aggregate_data2$average_by_interval == max(aggregate_data2$average_by_interval), ]$interval`
<br>

## Imputing missing values

Total rows with missing values : `r nrow(raw_data[is.na(raw_data$steps),])`
<br>
Missing values are replaced with the mean for the interval.
<br>

New histogram of the total number of steps taken each day:
<br>
``` {r}

#Impute missing values
clean_data <- raw_data
for (i in 1:nrow(clean_data)) {
  if (is.na(raw_data[i, ]$steps)) {
    clean_data[i, ]$steps <- aggregate_data2[(aggregate_data2$interval == clean_data[i,]$interval),]$average_by_interval
    
  }
}

#Aggregate the new data by day
aggregate_data3 <- aggregate(steps ~ date, clean_data, as.vector, na.action = na.pass)
aggregate_data3[, "total_daily"] <- 0
for (i in 1:nrow(aggregate_data3)) {
  aggregate_data3[i,]$total_daily <- sum(aggregate_data3[i,]$steps)
}


hist(aggregate_data3$total_daily, xlab = "Total Daily Steps", main = "Histogram of total daily steps")

```
<br>
New mean daily steps: `r format(mean(aggregate_data3$total_daily), digits = 2)`
<br>
New median daily steps: `r format(median(aggregate_data3$total_daily), digits = 2)`
<br>
As can be seen, the mean and median values do not change since NA values were ignored in the original computation and have simply been replaced with the mean in the clean data. 





## Are there differences in activity patterns between weekdays and weekends?


``` {r message = FALSE, warning=FALSE, fig.height = 12, fig.width = 12}

clean_data$day_type <- ""
levels(clean_data$day_type) <- c("Weekday", "Weekend")

weekend_days <- c("Saturday", "Sunday")
# Determine if the date represents a weekday or the weekend
for (i in 1:nrow(clean_data)) {

  testday <- weekdays(as.POSIXct(strptime(clean_data[i,]$date, "%Y-%m-%d"))) 
  clean_data[i,]$day_type <-   ifelse(testday %in% weekend_days, "Weekend", "Weekday")
}

intervals <- unique(clean_data$interval)

aggregate_data4 <- read.csv(text = "", col.name = c("average_steps", "interval", "day_type"), colClasses = "character")

for (i in intervals) {
  weekday_mean <- mean(clean_data[((clean_data$day_type == "Weekday") & (clean_data$interval == i)),]$steps)
  weekend_mean <- mean(clean_data[((clean_data$day_type == "Weekend") & (clean_data$interval == i)),]$steps)
  aggregate_data4 <- rbind(aggregate_data4, data.frame(average_steps = weekday_mean, interval = i, day_type = "Weekday"))
  aggregate_data4 <- rbind(aggregate_data4, data.frame(average_steps = weekend_mean, interval = i, day_type = "Weekend"))

}

par(mfrow = c(2, 1), mar = c(2.5, 4.1, 1.5, 2.1), omi = c(1,1,1,1))

plot(aggregate_data4[aggregate_data4$day_type == "Weekend",]$interval, aggregate_data4[aggregate_data4$day_type == "Weekend",]$average_steps, type = "l", xlim = c(0,2500), ylim = c(0, 250), xlab = NA, ylab = NA, main = "weekend", col = "blue" )

plot(aggregate_data4[aggregate_data4$day_type == "Weekday",]$interval, aggregate_data4[aggregate_data4$day_type == "Weekday",]$average_steps, type = "l", xlim = c(0,2500), ylim = c(0, 250), xlab = NA, ylab = NA, main = "weekday", col = "blue")

mtext("Interval", side = 1, outer = TRUE)
mtext("Number of steps", side = 2, outer = TRUE)



```
As can be seen from the plot above, the activity pattern for weekdays differs from that for the weekend.

